Gaze-dependent evidence accumulation predicts multi-alternative risky choice behaviour

### 2021-02-12 manuscript/manuscript_refs-numbered.tex
removed line numbers for preprint
separate page numbers for supplementary information

\usepackage{geometry}
\geometry{
    textwidth=18cm, textheight=23.4cm,marginratio={1:1,5:7}      % actual formatting
    % left=1cm, right=5cm, vmargin={2.5cm,3cm}, marginparwidth=5cm   % todo-note friendly formatting (https://tex.stackexchange.com/a/156328)
}

@@

\linenumbers  % uncomment to add line numbers

@@

Crucially, choices in these multi-alternative, multi-attribute settings pose a challenge to many traditional models of risky choice, including Expected Utility Theory (EU \autocite{vonneumann1947TheoryGamesEconomic}) and Prospect Theory \autocite{kahneman1979ProspectTheoryAnalysis,tversky1992AdvancesProspectTheory}, which obey basic axioms of rational choice like independence of irrelevant alternatives (IIA \autocite{luce1959IndividualChoiceBehavior}). Briefly, IIA asserts that the preference between two alternatives should not depend on other alternatives. At least three context effects – the attraction \autocite{huber1982AddingAsymmetricallyDominated}, compromise \autocite{simonson1989ChoiceBasedReasons}, and similarity \autocite{tversky1972EliminationAspectsTheory} effects – show that IIA is frequently violated: The attraction effect describes an increase in preference for an alternative following the addition of a similar, but slightly inferior alternative. The compromise effect describes an increase in preference for an alternative after a third alternative was added that makes it appear as intermediate. The similarity effect predicts that adding a third option that is similar to one of the original options, and equally appealing, will increase relative preference for the other, dissimilar alternative. While these effects are predominantly investigated using consumer goods, some studies also found them to affect choices between risky gambles \autocite{huber1982AddingAsymmetricallyDominated,mohr2017AttractionEffectRisky,soltani2012RangeNormalizationModelContextDependent,tversky1972EliminationAspectsTheory,wedell1991DistinguishingModelsContextually}.

@@

Expected Utility Theory (EU \autocite{vonneumann1947TheoryGamesEconomic}) assumes that choice behaviour can be described as maximization of expected subjective utility. We computed subjective utilities of option outcomes $m_i$ using a power function with one free parameter $\alpha$:

@@

Multialternative Decision Field Theory (MDFT\autocite{roe2001MultialternativeDecisionField}) is a dynamic connectionist model for multi-attribute, multi-alternative choice. MDFT can predict similarity, attraction and compromise effects. The core principle of MDFT is the calculation of valences $V(t)$ at each point in time $t$. Valences are determined as

@@

Individually best-fitting models were identified based on the Bayesian Information Criterion (BIC \autocite{schwarz1978EstimatingDimensionModel}), computed for each model $m$ as

@@

\subsection*{Statistical modelling}
We used Bayesian estimation (BEST \autocite{kruschke2013BayesianEstimationSupersedes,kruschke2014DoingBayesianData}) of differences, effect size $d$ and associated 95\% highest posterior density intervals (HDI$_{95}$) for all reported pairwise comparisons. Reported correlation coefficients and associated HDI$_{95}$ result from Bayesian correlation analyses \autocite{lee2013BayesianCognitiveModeling}. Regression estimates and HDI$_{95}$ result from Bayesian regression analyses implemented in PyMC3 \autocite{salvatier2016ProbabilisticProgrammingPython}.

@@

@@

\end{refsection} % (separate reference list for Supplement)

>>>>

\usepackage{geometry}
\geometry{textwidth=18cm, textheight=23.4cm,marginratio={1:1,5:7}}

@@

% \linenumbers  % uncomment to add line numbers

@@

Crucially, choices in these multi-alternative, multi-attribute settings pose a challenge to many traditional models of risky choice, including Expected Utility Theory (EU) \autocite{vonneumann1947TheoryGamesEconomic} and Prospect Theory \autocite{kahneman1979ProspectTheoryAnalysis,tversky1992AdvancesProspectTheory}, which obey basic axioms of rational choice like independence of irrelevant alternatives (IIA) \autocite{luce1959IndividualChoiceBehavior}. Briefly, IIA asserts that the preference between two alternatives should not depend on other alternatives. At least three context effects – the attraction \autocite{huber1982AddingAsymmetricallyDominated}, compromise \autocite{simonson1989ChoiceBasedReasons}, and similarity \autocite{tversky1972EliminationAspectsTheory} effects – show that IIA is frequently violated: The attraction effect describes an increase in preference for an alternative following the addition of a similar, but slightly inferior alternative. The compromise effect describes an increase in preference for an alternative after a third alternative was added that makes it appear as intermediate. The similarity effect predicts that adding a third option that is similar to one of the original options, and equally appealing, will increase relative preference for the other, dissimilar alternative. While these effects are predominantly investigated using consumer goods, some studies also found them to affect choices between risky gambles \autocite{huber1982AddingAsymmetricallyDominated,mohr2017AttractionEffectRisky,soltani2012RangeNormalizationModelContextDependent,tversky1972EliminationAspectsTheory,wedell1991DistinguishingModelsContextually}.

@@

Expected Utility Theory (EU) \autocite{vonneumann1947TheoryGamesEconomic} assumes that choice behaviour can be described as maximization of expected subjective utility. We computed subjective utilities of option outcomes $m_i$ using a power function with one free parameter $\alpha$:

@@

Multialternative Decision Field Theory (MDFT) \autocite{roe2001MultialternativeDecisionField} is a dynamic connectionist model for multi-attribute, multi-alternative choice. MDFT can predict similarity, attraction and compromise effects. The core principle of MDFT is the calculation of valences $V(t)$ at each point in time $t$. Valences are determined as

@@

Individually best-fitting models were identified based on the Bayesian Information Criterion (BIC) \autocite{schwarz1978EstimatingDimensionModel}, computed for each model $m$ as

@@

\subsection*{Statistical modelling}
We used Bayesian estimation (BEST) \autocite{kruschke2013BayesianEstimationSupersedes,kruschke2014DoingBayesianData} of differences, effect size $d$ and associated 95\% highest posterior density intervals (HDI$_{95}$) for all reported pairwise comparisons. Reported correlation coefficients and associated HDI$_{95}$ result from Bayesian correlation analyses \autocite{lee2013BayesianCognitiveModeling}. Regression estimates and HDI$_{95}$ result from Bayesian regression analyses implemented in PyMC3 \autocite{salvatier2016ProbabilisticProgrammingPython}.

@@

% Reset page counter and add "S"
\setcounter{page}{1}
\renewcommand{\thepage}{S\arabic{page}}

@@

\end{refsection} % (separate reference list for Supplement)
\renewcommand{\thepage}{S\arabic{page}}
